---
sidebar: false
toc: false
execute: 
  eval: false
---


# Spatial Transcriptomics

| Audience | Computational skills required| Duration |
:----------|:----------|:----------|
| Biologists | [Introduction to R](https://hbctraining.github.io/Intro-to-R-flipped/) | 4-session online workshop (~7.5 hours of trainer-led time)|

### Description

This repository has teaching materials for a hands-on **Introduction to Spatial Transcriptomics** workshop. **ADD DESCRIPTION HERE STILL** 

Working knowledge of R is required or completion of the [Introduction to R workshop](https://hbctraining.github.io/Intro-to-R-flipped/). 

::: callout-note
# Length of workshop
Please note that the schedule linked below assumes that learners will spend between 3-4 hours on reading through, and completing exercises from selected lessons between classes. The online component of the workshop focuses on more exercises and discussion/Q & A.

These materials were developed for a trainer-led workshop, but are also amenable to self-guided learning.
:::

### Learning Objectives

- LO 1
- LO 2
- LO 3
- LO 4


### Lessons
* [Workshop schedule (trainer-led learning)](schedule/schedule.qmd)
* [Self-learning](schedule/links-to-lessons.qmd)

### Installation Requirements

#### Applications
Download the most recent versions of R and RStudio for your laptop:

 - [R](http://lib.stat.cmu.edu/R/CRAN/) **(version 4.0.0 or above)**
 - [RStudio](https://www.rstudio.com/products/rstudio/download/#download)

#### Packages for R

::: callout-note
# Instructions for installing R packages

1. **Install the packages in the order listed below.**

2. **All the package names listed below are case sensitive.**

3. **Mac with Apple Silicon (M1/M2/etc.):**  
   Download and install the gfortran tool *before* installing your packages:  
   <https://mac.r-project.org/tools/gfortran-12.2-universal.pkg>

4. **Updating old packages:**  
   If R asks in the console:  
   `Update all/some/none? [a/s/n]:`  
   type **`a`** and press Enter to update any old packages.  
   _Note: Updating can take a while, so allow extra time before starting._

5. **“Binary version available but the source version is later” message:**  
   If you see:  
   `Do you want to install from sources the package which needs compilation? y/n`  
   type **`n`** (no) and press Enter.
:::

**(1)** Install the following packages:

**Please install them one-by-one as follows:**

```{r}
install.packages("tidyverse")
install.packages("Seurat")
install.packages("arrow")
install.packages("hdf5r")
install.packages("qs2")
install.packages("scales")
install.packages('leidenbase')
install.packages("remotes")
install.packages("R.utils")
install.packages("future")
install.packages("harmony")
install.package("devtools")

remotes::install_github("prabhakarlab/Banksy")
remotes::install_github('satijalab/seurat-wrappers')

devtools::install_github("dmcable/spacexr", 
                         build_vignettes = FALSE)

```

**(3)** Finally, please check that all the packages were installed successfully by **loading them one at a time** using the `library()` function.  

```{r}
library(package_1)
library(package_2)
library(package_3)
library(package_4)
library(package_5)
library(package_6)
library(package_7)
library(package_8)
library(package_9)
library(package_10)
library(package_11)
```

**(4)** Once all packages have been loaded, run sessionInfo().  

```{r}
sessionInfo()
```

---

### Citation

To cite material from this course in your publications, please use:

::: callout-important
# Citation
**Create Zenodo citation**
:::

A lot of time and effort went into the preparation of these materials. Citations help us understand the needs of the community, gain recognition for our work, and attract further funding to support our teaching activities. Thank you for citing this material if it helped you in your data analysis.
