---
title: "Quality Control"
description: |
  Focus on quality control strategies for Visium HD, using thresholds on genes, UMIs and mitochondrial content to filter low-quality spots. This lesson helps you detect technical artifacts and sample issues before downstream analyses.
author:
  - "Noor Sohail"
date: "2025-07-22"
categories: 
  - spatial transcriptomics
  - QC
  - Filtration
keywords:
  - Quality control
  - Mitochondrial ratio
  - UMI counts
license: "CC-BY-4.0"
editor_options: 
  markdown: 
    wrap: 72
---

```{r}
#| label: load_libraries_data
#| echo: false
# Load libraries and data
library(Seurat)
library(tidyverse)
seurat_merged <- qs2::qs_read("intermediate/03_seurat_merged.qs")
```

Approximate time: XX minutes

## Learning objectives 

In this lesson, we will:

- Construct quality control metrics and visually evaluate the quality of the data
- Apply appropriate filters to remove low quality cells
- Create a filtered seurat object

## Overview of lesson

In Visium HD data, the main challenge is in **delineating bins that are poor quality from bins containing reads from less complex cells.** If you expect a particular cell type in your dataset to be less transcriptionally active as compared other cell types in your dataset, the bins underneath this cell type will naturally have fewer detected genes and transcripts. However, having fewer detected genes and transcripts can also be a technical artifact and not a result of biological signal.

Here we will learn how to identify and set thresholds for filtration so that the dataset contains high-quality bins.

## Quality metrics

We will assess a variety of metrics to evaluate which bins are considered low/high quality. **We will apply very permissive filtering here** as it has been shown that low expression can be biologically meaningful for spatial context so we won’t be as stringent as we normally are with scRNA-seq. 

The metrics we will be using to filter low-quality bins from high-quality ones include:

- Cell counts
- UMI counts per cell
- Genes detected per cell
- Complexity (novelty score)
- Mitochondrial counts ratio

Some of these values we will calculate throughout this lesson and be added to our metadata dataframe.

```{r}
#| label: meta_view
#| eval: false
# Store metadata as variable meta
meta <- seurat_merged@meta.data
View(meta)
```

```{r}
#| label: tbl-meta_dt
#| tbl-cap: View `@meta.data` dataframe
#| echo: false
meta <- seurat_merged@meta.data
meta %>% 
  head(5) %>%
  knitr::kable()
```

::: {.callout-note collapse=true}
# "Nicer" spatial visualizations
For "nicer" (subjective) plotting with `SpatialFeaturePlot()` and `SpatialDimPlot()`, we will add some extra parameters to gain a clearer image beyond the default plot (shown here):

```{r}
#| label: fig-nCount_spatial_default
#| fig-cap: Default `SpatialFeaturePlot()` visualization parameters.
SpatialFeaturePlot(seurat_merged, 
                   "nCount_Spatial.008um",
                   pt.size.factor = 15)
```

Therefore for the rest of this lesson, we will be using the following arguments:

- `pt.size.factor = 15`: to clearly see each bin on the slide
- `image.alpha = 0`: to remove the H&E stained image in the background of the image
- `max.cutoff` and `min.cutoff`: to not allow the color scale to be driven by smaller populations of cells with high/low values
:::

We will be using a variety of visualization methods:

- Looking at the values of each bin on the spatial slide .
- Distribution of values before and after filtration as a density plot.

### Number of bins

First, we can see how many bins we have per sample. This is a number we should keep in the back of our mind throughout the analysis because it will help us understand the distribution of our spots. For spatial datasets, the number of bins is going to correspond directly to the bin size chosen as well as the bin size selected.

```{r}
#| label: fig-n_cells
#| fig-cap: Number of bins in the dataset, split by sample.
# Barplot number of bins per sample
ggplot(meta) +
  geom_bar(aes(x = orig.ident, fill = orig.ident),
           color = "black") +
  geom_text(aes(x = orig.ident, label=after_stat(count)), 
            stat='count', vjust=-1) +
  theme_classic()
```


### UMI counts (transcripts) per bin

The `nCounts` is the number of unique transcripts (UMIs) detected per bin. Oftentimes, these values will correspond with how transcriptionally active a cell may be, which is typically defined by their celltype. For example, tumor cells will have very high UMI counts.

In the density plots, we expect to see a bimodal distribution. One peak should represent bins containing lower-quality bins with fewer UMIs and another peak representing healthy bins with more UMIs. Ideally, the peak representing lower-quality and dying cells is small and the peak representing healthy cells is large.

These numbers may be lower than what we would expect for scRNA-seq datasets due to the small size of the bins.

::: {.panel-tabset}

#### Spatial overlay

```{r}
#| label: fig-nCount_spatial
#| fig-cap: Number of UMIs overlaid over spatial slide.
SpatialFeaturePlot(seurat_merged, 
                   "nCount_Spatial.008um",
                   pt.size.factor = 15,
                   image.alpha = 0,
                   max.cutoff = "q90")
```

#### Before filtration density

```{r}
#| label: fig-nCount_density
#| fig-cap: Number of UMIs density.
ggplot(meta) +
  geom_density(aes(x = nCount_Spatial.008um, fill = orig.ident),
               alpha = 0.4,
               color = "black") +
  geom_vline(xintercept = 30, color = "pink") +
  geom_vline(xintercept = 10, color = "lightblue") +
  scale_x_log10() +
  theme_classic()
```


#### After filtration density

```{r}
#| label: fig-nCount_density_filt
#| fig-cap: Number of UMIs density after filtration.
meta_filt <- subset(meta,
  ((orig.ident == "P5CRC") & (nCount_Spatial.008um > 30)) |
  ((orig.ident == "P5NAT") & (nCount_Spatial.008um > 10)))

ggplot(meta_filt) +
  geom_density(aes(x = nCount_Spatial.008um,
                     fill = orig.ident),
                 alpha = 0.4,
                 color = "black") +
  geom_vline(xintercept = 30, color = "pink") +
  geom_vline(xintercept = 10, color = "lightblue") +
  scale_x_log10() +
  theme_classic()
```

:::

::: {.callout-note collapse=true}
# Sample specific values
A commonly asked question is if you have to use the same threshold values across all your samples. We recommend that you follow what the data is telling you and do it on a per sample basis. Ultimately the end goal is to retain **high-quality bins**, even if that means using different values for each sample.
:::

### Genes detected per bin

This is the number of genes (`nFeatures`) detected per bin, or the number of genes that have a non-zero value in a bin. We have similar expectations for gene detection as we did for number of UMIs in terms of the distribution of values.

When we look at the spatial slide, we can already beging to see patterns of expression (both with genes and UMIs) that will be helpful in understanding our dataset better.

::: {.panel-tabset}

#### Spatial overlay

```{r}
#| label: fig-nFeature_spatial
#| fig-cap: Number of features overlaid over spatial slide.
SpatialFeaturePlot(seurat_merged, 
                   "nFeature_Spatial.008um",
                   pt.size.factor = 15,
                   image.alpha = 0,
                   max.cutoff = "q90")
```

#### Before filtration density

```{r}
#| label: fig-nFeature_density
#| fig-cap: Number of features density.
ggplot(meta) +
  geom_density(aes(x = nFeature_Spatial.008um,
                     fill = orig.ident),
                 alpha = 0.4,
                 color = "black") +
  geom_vline(xintercept = 30, color = "pink") +
  geom_vline(xintercept = 10, color = "lightblue") +
  scale_x_log10() +
  theme_classic()
```


#### After filtration density

```{r}
#| label: fig-nFeature_density_filt
#| fig-cap: Number of features density after filtration.
meta_filt <- subset(meta,
  ((orig.ident == "P5CRC") & (nFeature_Spatial.008um > 30)) |
  ((orig.ident == "P5NAT") & (nFeature_Spatial.008um > 10)))

ggplot(meta_filt) +
  geom_density(aes(x = nFeature_Spatial.008um,
                     fill = orig.ident),
                 alpha = 0.4,
                 color = "black") +
  geom_vline(xintercept = 30, color = "pink") +
  geom_vline(xintercept = 10, color = "lightblue") +
  scale_x_log10() +
  theme_classic()
```

:::

### Complexity (novelty) score

Sometimes there may be bins with high `nCounts` (UMIs) and low `nFeatures` (genes). This would indicate that few genes were sequenced many times over. We consider these instances to be cases of "low complexity" where we are getting high expression from few genes. Some celltypes, such as red blood cells, are known for such behavior. However, we should be cautious when interpreting these results, as contamination or technical artifact could also contribute to this. Generally, we expect the complexity score to be above 0.80 for good-quality bins.

The novelty score is computed as a ratio of genes to UMIs, as shown below:

$$
\text{Complexity Score}
= \frac{\log_{10}(\text{Number of Genes})}{\log_{10}(\text{Number of UMIs})}
$$

Which we can now calculate using R and store in our `@meta.data`:

```{r}
#| label: calc_log10GenesPerUMI
# Add number of genes per UMI for each cell to metadata
seurat_merged$log10GenesPerUMI <- log10(seurat_merged$nFeature_Spatial.008um) / 
                                  log10(seurat_merged$nCount_Spatial.008um)

# Turn NA values into 0 for now
seurat_merged$log10GenesPerUMI[is.na(seurat_merged$log10GenesPerUMI)] <- 0
```

::: {.panel-tabset}

#### Spatial overlay

```{r}
#| label: fig-log10GenesPerUMI_spatial
#| fig-cap: Complexity score overlaid over spatial slide.
SpatialFeaturePlot(seurat_merged, 
                   "log10GenesPerUMI",
                   pt.size.factor = 17,
                   image.alpha = 0,
                   min.cutoff = "q10")
```

#### Before filtration density

```{r}
#| label: fig-log10GenesPerUMI_density
#| fig-cap: Complexity score density.
meta <- seurat_merged@meta.data
ggplot(meta) +
  geom_density(aes(x = log10GenesPerUMI,
                   fill = orig.ident),
                 alpha = 0.4,
                 color = "black") +
  geom_vline(xintercept = 0.80) +
  theme_classic()
```


#### After filtration density

```{r}
#| label: fig-log10GenesPerUMI_density_filt
#| fig-cap: Complexity score density after filtration.
meta_filt <- subset(meta, log10GenesPerUMI > 0.80)
ggplot(meta_filt) +
  geom_density(aes(x = log10GenesPerUMI,
                   fill = orig.ident),
                 alpha = 0.4,
                 color = "black") +
  geom_vline(xintercept = 0.80) +
  theme_classic()
```

:::

### Mitochondrial counts ratio

During sequencing, we do not only measure the levels of expression the nuclear genome - we are also capturing the mitochondrial genome! High levels of mitochondrial expression is a sign that the cell is dying or dead. Therefore, we can calculate the proportion of reads (UMIs) that come from the mitochondria out of all the transcripts in a cell as another metric.

 We define poor-quality samples for mitochondrial counts as bins which surpass the 0.25 mitochondrial ratio threshold, unless of course you are expecting this in your sample. This ratio is computed as:

$$
\text{Mitochondrial Ratio} = 
\frac{\text{Number of reads aligning to mitochondrial genes}}
     {\text{Total reads}}
$$

::: {.callout-note collapse=true}
# Think about your biological question
While using a baseline score of 0.25 is an acceptable threshold for removing high mitochondrial content cells, it is important to always go back to your original biological question. What samples are you working with? Do you expect there to be high values of mitochondria due to your experimental condition?

For example, if you were studying renal oncocytomas would you make this same choice? This disease is characterized as having [aberrantly high mitochondrial expression](https://www.nature.com/articles/modpathol2015101), would it make sense to remove cells with high mitochondrial ratio?
:::


```{r}
#| label: calc_mitoRatio
# Compute percent mito ratio
seurat_merged$mitoRatio <- PercentageFeatureSet(object = seurat_merged, 
                                                pattern = "^MT-")
seurat_merged$mitoRatio <- seurat_merged@meta.data$mitoRatio / 100

# Turn NA values into 100 for now
seurat_merged$mitoRatio[is.na(seurat_merged$mitoRatio)] <- 1.00
```

::: {.panel-tabset}

#### Spatial overlay

```{r}
#| label: fig-mitoRatio_spatial
#| fig-cap: Mitochondrial ratio overlaid over spatial slide.
SpatialFeaturePlot(seurat_merged, 
                   "mitoRatio",
                   pt.size.factor = 15,
                   image.alpha = 0,
                   max.cutoff = "q90")
```

#### Before filtration density

```{r}
#| label: fig-mitoRatio_density
#| fig-cap: Mitochondrial ratio density.
meta <- seurat_merged@meta.data
ggplot(meta) +
  geom_density(aes(x = mitoRatio,
                   fill = orig.ident),
                 alpha = 0.4,
                 color = "black") +
  geom_vline(xintercept = 0.25) +
  theme_classic()
```


#### After filtration density

```{r}
#| label: fig-mitoRatio_density_filt
#| fig-cap: Mitochondrial ratio density after filtration.
meta_filt <- subset(meta, mitoRatio < 0.25)

ggplot(meta_filt) +
  geom_density(aes(x = mitoRatio,
                   fill = orig.ident),
                 alpha = 0.4,
                 color = "black") +
  geom_vline(xintercept = 0.25) +
  theme_classic()
```

:::

:::{.callout-tip}
# [**Exercise 1**](05_quality_control-Answer_key.qmd#exercise-1)
1. Do you notice a pattern in cells in regards to the number of UMIs and features? Make a `geom_point` plots to compare these values on a per-cell basis and color each point by the mitochondrial ratio following the structure provided here:

```{r}
#| label: exercise_geom_point
#| eval: false
# Structure for making geom_point plot
# Fill in values to answer the question
seurat_merged@meta.data %>%
  # Sorting by mitoRatio to make high scores appear on top of the plot
  arrange(mitoRatio) %>%
  ggplot() +
  geom_point(aes(x = ???, 
                 y = ???,
                 color = ???),
             size = 0.5) +
  # Setting limits so that outliers don't determine scale of the plot
  ylim(0, 3500) + xlim(0, 3500) +
  theme_bw()
```

:::

## Filtration

We will apply very minimal filtering here. It has been shown that low expression can be biologically meaningful for spatial context so we won’t be as stringent as we normally are with scRNA-seq.

```{r}
#| label: filter_cells
# Per-sample nCount thresholds
seurat_filtered <- subset(seurat_merged,
  ((orig.ident == "P5CRC") & (nCount_Spatial.008um > 30)) |
  ((orig.ident == "P5NAT") & (nCount_Spatial.008um > 10)))

# Per-sample nFeature thresholds
seurat_filtered <- subset(seurat_filtered,
  ((orig.ident == "P5CRC") & (nFeature_Spatial.008um > 30)) |
  ((orig.ident == "P5NAT") & (nFeature_Spatial.008um > 10)))
  
# Global thresholds for mitochondrial ratio and complexity
seurat_filtered <- subset(seurat_filtered, mitoRatio < 0.25)
seurat_filtered <- subset(seurat_filtered, log10GenesPerUMI > 0.80)

# Print seurat object after filtration
seurat_filtered
```


::: callout-warning
After subsetting, you may get the following warning message:

```
Warning: Not validating Centroids objects 
Warning: Not validating FOV objects
Warning: Not validating FOV objects
Warning: Not validating FOV objects
Warning: Not validating Seurat objects
```

This warning message can be ignored as it is Seurat internalling checking bin barcodes against the image. After every `subset` step this message may appear, but can ignored as ultimately the subsetting is still accomplished.

:::


How many cells did we remove in this filtration process?

```{r}
#| label: calc_ncells_filtered
# Differences of cells before and after filtration
ncol(seurat_merged) - ncol(seurat_filtered)
```

:::{.callout-tip}
# [**Exercise 2**](05_quality_control-Answer_key.qmd#exercise-2)
2. How many bins do we have per sample after this filtration step?
:::

## Visualizing counts data

We can visualize the number of UMIs and gene counts per bin, both as a distribution and layered on top of the tissue image. 

### Violin plots

Let’s start with a violin plot to look at the distribution of UMI counts and gene counts. The input is our post-filtered dataset.

```{r}
#| label: fig-vln_ncount_nfeature
#| fig-cap: Violin plot of nCounts and nFeatures after filtration.
p_ncount <- VlnPlot(seurat_filtered, 
                    features = "nCount_Spatial.008um", 
                    pt.size = 0, group.by = 'orig.ident') +
  NoLegend()

p_nfeats <- VlnPlot(seurat_filtered, 
                    features = "nFeature_Spatial.008um", 
                    pt.size = 0, group.by = 'orig.ident') + 
  NoLegend()

p_ncount | p_nfeats
```

We see that both distributions have a similar peak but the nUMI distribution has a much longer tail. This is expected, because while the small physical size of the bins means that most genes will be detected only once or twice, a minority of bins under very transcriptionally active cells may exhibit multiple transcripts of the same gene.

### Spatial overlay

Next, we can look at the same metrics and the distribution on the actual image itself. Note that many spots have very few counts, in part due to low cellular density or cell types with low complexity in certain tissue regions.

```{r}
#| label: fig-nFeature_nCount_spatial
#| fig-cap: Number of features and counts overlaid over spatial slide.
#| fig-width: 10
#| fig-height: 8
SpatialFeaturePlot(seurat_filtered, 
                   c("nFeature_Spatial.008um", 
                     "nCount_Spatial.008um"),
                   pt.size.factor = 16,
                   image.alpha = 0)
```


## Save!

Now is a great spot to save our `seurat_filtered` object as we have finished filtering.

```{r}
#| label: save_seurat_filtered
#| eval: false
# Save Seurat object
saveRDS(seurat_filtered, "data/seurat_filtered.RDS")
```

```{r}
#| label: save_seurat_filtered_qs
#| eval: false
#| echo: false
# Save Seurat object
qs2::qs_save(seurat_filtered, "intermediate/05_seurat_filtered.qs")
```

***

[Next Lesson >>](06_normalization_and_sketch.qmd)

[Back to Schedule](../schedule/schedule.qmd)