---
title: "Normalization and Sketch Downsampling - Answer Key"
author:
  - Noor Sohail
date: "2025-07-22"
license: "CC-BY-4.0"
editor_options: 
  markdown: 
    wrap: 72
---

```{r}
#| label: load_libraries_data
#| echo: false
# Load libraries and data
library(Seurat)
library(tidyverse)

seurat_filtered <- qs2::qs_read("intermediate/05_seurat_filtered.qs")
seurat_filtered <- UpdateSeuratObject(seurat_filtered)
seurat_sketch <- NormalizeData(seurat_filtered)
```

# Exercise 1

1. Using [GeneCards](https://www.genecards.org/), look up one of the top genes and read more about its role and how it could possible relate to CRC.

| Gene   | Genecard description                                                                                                                                          | Marker                          |
|--------|---------------------------------------------------------------------------------------------------------------------------------------------------------------|---------------------------------|
| IGHG1, IGLC1, IGLC7, IGHG3, IGHA1 | membrane-bound immunoglobulins serve as receptors which, upon binding of a specific antigen, trigger the clonal expansion and differentiation of B lymphocytes into immunoglobulins-secreting plasma cells | B cell -> plasma cell marker    |
| IGHM   | antigen recognition molecules of B cells                                                                                                                      | B cell marker                   |
| SST    | Somatostatin also affects rates of neurotransmission in the central nervous system and proliferation of both normal and tumorigenic cells.                   | ?                               |
| INSL5  | May have a role in gut contractility                                                                                                                          | ?                               |
| CHGA   | It is found in secretory vesicles of neurons and endocrine cells                                                                                              | Neuron + endocrine celltype marker |
| CXCL8  | ?                                                                                                                                                             | ?                               |
| GCG    | secreted from gut endocrine cells and promote nutrient absorption                                                                                            | ?                               |
| HBA2   | ?                                                                                                                                                             | ?                               |
| VIP    | It stimulates myocardial contractility, causes vasodilation, increases glycogenolysis, lowers arterial blood pressure and relaxes the smooth muscle of trachea, stomach and gall bladder | ?                               |
| MMP12  | Macrophage Metalloelastase                                                                                                                                    | Macrophage marker               |
| PYY    | secreted by endocrine cells in the gut                                                                                                                        | Endocrine marker                |

You might not always be able to identify what the exact role of a gene is at this point. However, now we have a better understanding of what genes are going to be driving the downstream steps of our analysis.

2. Visualize one of the top variable genes on the spatial slide to see if expression varies across the dataset.

```{r}
SpatialFeaturePlot(seurat_sketch,
                   features = "IGHG1",
                   image.alpha = 0,
                   pt.size.factor = 15)
```

We can see that the expression of IGHG1 appears to have patches of bins with very high expression, whereas in other areas it has near 0 expression. This confirms that it is a gene that is expressed variably throughout the dataset.

```{r}
SpatialFeaturePlot(seurat_sketch,
                   features = "IGHM",
                   image.alpha = 0,
                   pt.size.factor = 15)
```

Similarly, we can see that gene IGHM, which is a B-cell marker genes, is expressed selectively throughout the dataset.