---
title: "Normalization and Sketch Downsampling - Answer Key"
author:
  - Noor Sohail
date: "2025-07-22"
license: "CC-BY-4.0"
editor_options: 
  markdown: 
    wrap: 72
---

```{r}
#| label: load_libraries_data
#| echo: false
# Load libraries and data
library(Seurat)
library(tidyverse)

seurat_filtered <- qs::qread("intermediate/05_seurat_filtered.qs")
seurat_sketch <- NormalizeData(seurat_filtered)
```

# Exercise 1

1. Using [GeneCards](https://www.genecards.org/), look up one of the top genes and read more about its role and how it could possible relate to CRC.

::: callout-warning
TODO
:::

2. Visualize one of the top variable genes on the spatial slide to see if expression varies across the dataset.

```{r}
SpatialFeaturePlot(seurat_sketch,
                   features = "IGHG1",
                   image.alpha = 0,
                   pt.size.factor = 15)
```

We can see that the expression of IGHG1 appears to have patches of bins with very high expression, whereas in other areas it has near 0 expression. This confirms that it is a gene that is expressed variably throughout the dataset.

```{r}
SpatialFeaturePlot(seurat_sketch,
                   features = "IGHM",
                   image.alpha = 0,
                   pt.size.factor = 15)
```

Similarly, we can see that gene IGHM, which is a B-cell marker genes, is expressed selectively throughout the dataset.