---
title: "Deconvolution"
authors: "Will Gammerdinger, Noor Sohail, Zhu Zhuo, James Billingsley, Shannan Ho Sui"
date: "Tuesday, July 22, 2025"
editor_options: 
  markdown: 
    wrap: 72
---

Approximate time: XY minutes

# Learning Objectives

-   LO 1
-   LO 2
-   LO 3

# What is Deconvolution?

The single-digit micron resolution is a big technological improvement over Visium's original ∼55μm spots. Because a single spot can include several cell types, the dataset is not straightforward to annotate. As a result, many methods **model each spot as a mixture**, assigning proportion values for each cell type within that spot. Each spot is represented as a piechart showing proportions of the cell type composition.

![[Ma, Y., et al. Nat Biotechnol (2022)](https://www.nature.com/articles/s41587-022-01273-7), Figure 1](../img/card_deconvolution.png){width="50%"}

While Visium HD reduces bin size to approach single-cell resolution, it is important to note that sequencing-based spatial transcriptomics captures RNA from a defined spatial area rather than from isolated cells. As a result, transcripts from multiple cells overlapping a bin can be pooled and sequenced together. This means that even small bins may contain mixed cellular signals, which has important implications for downstream analyses such as differential gene expression.


::: callout-tip
# [**Exercise 1**](09_integration_Answer-key.qmd#exercise-1)

Exercise 1
:::

# Robust Cell Type Deconvolution 

```{r}
library(Seurat)
library(tidyverse)
library(spacexr)

seurat_banksy <- readRDS("data/seurat_banksy_2.RDS")
```

To perform accurate annotation of cell types, we must also take into consideration that our 16µm x 16µm bins may contain multiple cells. The method [Robust Cell Type Deconvolution](https://www.nature.com/articles/s41587-021-00830-w) (RCTD) has been shown to accurately annotate spatial data from a variety of technologies while taking into consideration that a single bin may exhibit multiple cell type profiles.

We are going to annotate the slides one a time. There are extensions of the RCTD package in order to take into account [replicates and multiple samples](https://raw.githack.com/dmcable/spacexr/master/vignettes/replicates.html) which we will not use at this time.

RCTD takes a cell-type-annotated scRNA-seq dataset as a reference and a spatial dataset as a query. For our reference, we will use a paired scRNA-seq FLEX dataset that was generated from the same patients.

## Load Reference Dataset

The two most important pieces of information necessary to create a RCTD `Reference()` object are counts and celltype annotations. We can take a quick look at the different celltypes that we are going to annotate our query object with. For this reference, these celltypes values are stored in the columns `Level1` and `Level2` - with varying levels of granularity for the annotation.

TODO: have to downsample referenece dataset

::: {.callout-note collapse="true"}
The reference dataset was generated by downloading two files with bash:

- Count matrix (`.h5` file)
- Metadata csv

```{bash}
#| eval: false
curl -O https://cf.10xgenomics.com/samples/cell-exp/8.0.0/HumanColonCancer_Flex_Multiplex/HumanColonCancer_Flex_Multiplex_count_filtered_feature_bc_matrix.h5
wget "https://github.com/10XGenomics/HumanColonCancer_VisiumHD/raw/refs/heads/main/MetaData/SingleCell_MetaData.csv.gz"
```

Then in R, both files were loaded in and put together to create a seurat object. This is the same object that we will be loading in the next step of this lesson:

```{r}
#| eval: false
counts <- Read10X_h5("HumanColonCancer_Flex_Multiplex_count_filtered_feature_bc_matrix.h5")
meta <- read.csv("SingleCell_MetaData.csv")
rownames(meta) <- meta$Barcode

seurat <- CreateSeuratObject(counts = counts,
                             meta.data = meta,
                             project = "CRC FLEX")

# Grab UMAP coordinates and put them in a reduction
umap_coords <- as.matrix(seurat@meta.data[, c("UMAP1", "UMAP2")])

# Create a DimReduc and store it in the object as "umap"
seurat[["umap"]] <- CreateDimReducObject(embeddings = umap_coords,
                                         key = "UMAP_",
                                         assay = DefaultAssay(seurat))


# Remove cells that did not pass QC
seurat <- subset(seurat, subset = (QCFilter == "Keep"))
```

:::

So first, let us load in the reference seurat object that is provided:

```{r}
seurat_ref <- readRDS("data/crc_flex_ref.RDS")
seurat_ref
```

We know that there are two different levels of celltype annotations which we can begin investigating.

```{r}
#| fig-width: 15
# UMAP level 1
p1 <- DimPlot(seurat_ref, 
              group.by = "Level1", 
              label = TRUE) +
  ggtitle("CRC Reference: Level 1")

# UMAP level 2
p2 <- DimPlot(seurat_ref, 
              group.by = "Level2", 
              label = TRUE) +
  ggtitle("CRC Reference: Level 2")

p1 | p2
```

```{r}
Idents(seurat_ref) <- "Level2"
seurat_ref <- subset(x = seurat_ref, downsample = 500)
seurat_ref
```

::: callout-tip
# [**Exercise 2**](09_integration_Answer-key.qmd#exercise-2)

What are the differences between `Level1` and `Level2`. Calculate how many cells belong to each celltype and identify which ones are most abundant in the reference dataset.

```{r}
# Plot number of bins per Level1 celltype
ggplot(seurat_ref@meta.data, 
       aes(x = Level1, fill = Level1)) +
  geom_bar() +
  stat_count(aes(label = after_stat(count)),
             geom  = "text",vjust = -0.3) +
  theme_bw() + NoLegend() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  ggtitle("CRC Reference: Level1")
```

```{r}
#| fig-width: 10
# Plot number of bins per Level2 celltype
ggplot(seurat_ref@meta.data, 
       aes(x = Level2, fill = Level2)) +
  geom_bar() +
  stat_count(aes(label = after_stat(count)),
             geom  = "text",vjust = -0.3) +
  theme_bw() + NoLegend() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  ggtitle("CRC Reference: Level2")
```
:::


Now that we better understand the reference, we are going to make a RCTD object out of it. This time we use the `Reference()` function and additionally supply `cluster` (celltype) information to transfer to our query. We supply the following three parameters:

- `counts` = raw counts matrix
- `cell_types` = celltype annotation
- `nUMI` = total number of counts per cell


```{r}
# Raw counts of reference dataset
counts <- seurat_ref[["RNA"]]$counts

# Celltype annotation of reference dataset
cluster <- seurat_ref$Level2
cluster <- as.factor(cluster)

# Total counts of each cell
nUMI <- seurat_ref$nCount_RNA

# Create the RCTD reference object
reference <- Reference(counts = counts, 
                       cell_types = cluster, 
                       nUMI = nUMI)
```

## Prepare Query

To begin, we are going to split our seurat object into a single sample.

```{r}
crc <- subset(seurat_banksy, subset = (orig.ident == "P5CRC"))
crc
```

```{r}
# Create sketch of CRC subset
DefaultAssay(crc) <- 'Spatial.016um'

# HVGs are required input for SketchData
crc <- FindVariableFeatures(object = crc,
                            assay = "Spatial.016um")
crc <- ScaleData(crc, assay = "Spatial.016um")
crc <- RunPCA(crc, assay = "Spatial.016um", reduction.name = "pca_crc")
```

Deconvolution is a very computationally challenging task. So we can once again make use the `sketch` workflow to downsample our dataset in a careful manner.

```{r}
crc <- SketchData(
  object = crc,
  ncells = 5000,
  method = "LeverageScore",
  sketched.assay = "sketch_crc",
  var.name = "leverage.score_crc")

# We need the PCA to later project back onto the entire dataset
crc <- FindVariableFeatures(crc)
crc <- ScaleData(crc)
crc <- RunPCA(crc, reduction.name = "pca.sketch_crc")
```

RCTD requires a unique data structure (similar to Seurat) as input to run. So we create our query object with `SpatialRNA()` by suppling the spatial coordinates and raw counts for the bins.

```{r}
DefaultAssay(crc) <- "sketch_crc"

# Grab barcodes for sketched bins
counts_sketch <- crc[["sketch_crc"]]$counts
cells_sketch <- colnames(crc[["sketch_crc"]])

# Spatial coordinates for each bin
coords <- GetTissueCoordinates(crc)[cells_sketch, 1:2]

# Create the RCTD query object
query <- SpatialRNA(coords = coords, 
                    counts = counts_sketch, 
                    nUMI = colSums(counts_sketch))
```


## Run RCTD

Start: 11:46
End: 1:27

Start: 1:45
End: 3:27

```{r}
# Run RCTD
RCTD <- create.RCTD(spatialRNA = query, 
                    reference = reference, 
                    max_cores = 6)
RCTD <- run.RCTD(RCTD, doublet_mode = "doublet") # 
```

## RCTD Results

The resultant dataframe from RCTD will contain the following columns according to the [documentation](https://mirror.accum.se/mirror/bioconductor.org/packages/devel/bioc/vignettes/spacexr/inst/doc/rctd-tutorial.html):

-   `spot_class`, a factor variable representing RCTD's classification in doublet mode:
    -   "singlet" (1 cell type on pixel)
    -   "doublet_certain" (2 cell types on pixel)
    -   "doublet_uncertain" (2 cell types on pixel, but only confident of 1)
    -   "reject" (no prediction given for pixel)
-   `first_type` column gives the first cell type predicted on the bead (for all spot_class conditions except "reject").
-   `second_type` column gives the second cell type predicted on the bead for doublet spot_class conditions (not a confident prediction for "doublet_uncertain").


The results of deconvolution are stored in the `results$results_df` slot of the `RCTD` object:

```{r}
df_rctd_results <- RCTD@results$results_df
df_rctd_results %>% head()
```

# RCTD Results in Seurat

We want to add the relevant information into our seurat object using the `AddMetaData()` function.

```{r}
crc <- AddMetaData(crc, 
                   metadata = RCTD@results$results_df)
```

```{r}
table(crc$second_type)
```

## Project Back to Full Dataset

```{r}
# Project back first_type to new column called full_first_type
crc <- ProjectData(
  object = crc,
  assay = "Spatial.016um",
  sketched.assay = "sketch_crc",
  full.reduction = "pca_crc",
  sketched.reduction = "pca.sketch_crc",
  dims = 1:30,
  refdata = list(full_first_type = "first_type",
                 full_second_type = "second_type"))
```

## Visualize New Annotation

```{r}
SpatialDimPlot(crc,
               group.by = "full_first_type")
```

::: callout-tip
# [**Exercise 3**](09_integration_Answer-key.qmd#exercise-3)

Exercise 3
:::