---
title: "Integration - Answer Key"
author:
  - Noor Sohail
date: "2026-04-19"
license: "CC-BY-4.0"
editor_options: 
  markdown: 
    wrap: 72
---

```{r}
#| label: load_libraries_data
#| echo: false
# Load libraries and data
library(Seurat)
library(tidyverse)
seurat_processed <- qs2::qs_read("intermediate/08_seurat_processed.qs")

```

# Exercise 1

1. Do you see a clear split based on sample in the UMAP and PCA? What could this indicate about the presence of batch effects in our data?

```{r}
#| label: fig-pca_umap_by_sample
#| fig-cap: PCA and UMAP colored by sample.
#| fig-width: 10
#| echo: false
p1 <- DimPlot(seurat_processed,
              group.by="orig.ident",
              reduction = "full.pca.sketch") +
  NoLegend() + ggtitle("PCA (Projected)")

p2 <- DimPlot(seurat_processed,
              group.by = "orig.ident",
              reduction = "full.umap.sketch") +
  ggtitle("UMAP (Projected)")

p1 | p2
```

When we look at the UMAP and PCA plot without any other pieces of information it does seem like there is a batch effect. In the PCA space, we can see in the top left-hand corner of PC space that the bins predominantly belong to `P5CRC`. Similarly in the UMAP space, there is a cluster of bins on the right-hand side that belong solely to `P5CRC` and another at the top of the UMAP.

Based solely on this visualization and no additional information, it may seem like integration may be necessary - though the other bins in the dataset do appear to overlay quite well across both datasets.

# Exercise 2

2. Is there a biological reason we see clusters that are dominated by `P5CRC` bins? What cell type do you think these clusters may represent?

When we look at the celltypes we anticipate being in our dataset, there is one population that stands outs - `Tumor cells`.

The `P5NAT` sample is "normal adjacent tissue" which indicates that we should not see any tumor cells in this sample. Conversely, we are anticipating cancerous cells in the `P5CRC` dataset. So before we make a final decision on integration, we should first see if the previously identified populations could possible be tumor cells.

***

[Back to Lesson >>](09_integration.qmd)

[Back to Schedule](../schedule/schedule.qmd)