---
title: "Deconvolution"
description: |
  Write a description of the lesson here. 
author:
  - Noor Sohail
date: "2025-07-22"
categories: 
  - category_1
  - category_2
  - category_3
  - category_4
keywords:
  - keyword_1
  - keyword_2
  - keyword_3
  - keyword_4
  - keyword_5
  - keyword_6
license: "CC-BY-4.0"
editor_options: 
  markdown: 
    wrap: 72
---

```{r}
#| label: load_libraries_data
#| echo: false
# Load libraries and data
library(Seurat)
library(tidyverse)
library(spacexr)
seurat_banksy <- qs::qread("intermediate/09_seurat_banksy_2.qs")
```

Approximate time: XX minutes

## Learning objectives 

In this lesson, we will:

- Learning Objective 1
- Learning Objective 2
- Learning Objective 3

## Overview of lesson

When doing XYZ...

## What is Deconvolution?

The single-digit micron resolution is a big technological improvement over Visium's original ∼55μm spots. Because a single spot can include several cell types, the dataset is not straightforward to annotate. As a result, many methods **model each spot as a mixture**, assigning proportion values for each cell type within that spot. Each spot is represented as a piechart showing proportions of the cell type composition.

![[Ma, Y., et al. Nat Biotechnol (2022)](https://www.nature.com/articles/s41587-022-01273-7), Figure 1](../img/card_deconvolution.png){width="50%"}

While Visium HD reduces bin size to approach single-cell resolution, it is important to note that sequencing-based spatial transcriptomics captures RNA from a defined spatial area rather than from isolated cells. As a result, transcripts from multiple cells overlapping a bin can be pooled and sequenced together. This means that even small bins may contain mixed cellular signals, which has important implications for downstream analyses such as differential gene expression.


## Robust Cell Type Deconvolution 

```{r}
#| eval: false
library(Seurat)
library(tidyverse)
library(spacexr)
library(qs)

seurat_banksy <- readRDS("data/seurat_banksy_2.RDS")
```

To perform accurate annotation of cell types, we must also take into consideration that our 16µm x 16µm bins may contain multiple cells. The method [Robust Cell Type Deconvolution](https://www.nature.com/articles/s41587-021-00830-w) (RCTD) has been shown to accurately annotate spatial data from a variety of technologies while taking into consideration that a single bin may exhibit multiple cell type profiles.

We are going to annotate the slides one a time. There are extensions of the RCTD package in order to take into account [replicates and multiple samples](https://raw.githack.com/dmcable/spacexr/master/vignettes/replicates.html) which we will not use at this time.

RCTD takes a cell-type-annotated scRNA-seq dataset as a reference and a spatial dataset as a query. For our reference, we will use a paired scRNA-seq FLEX dataset that was generated from the same patients.


## Reference Dataset

The two most important pieces of information necessary to create a RCTD `Reference()` object are counts and celltype annotations. We can take a quick look at the different celltypes that we are going to annotate our query object with. For this reference, these celltypes values are stored in the columns `Level1` and `Level2` - with varying levels of granularity for the annotation.

::: callout-warning
TODO: have to downsample referenece dataset
:::

::: {.callout-note collapse="true"}
The reference dataset was generated by downloading two files with bash:

- Count matrix (`.h5` file)
- Metadata csv

```{bash}
#| eval: false
curl -O https://cf.10xgenomics.com/samples/cell-exp/8.0.0/HumanColonCancer_Flex_Multiplex/HumanColonCancer_Flex_Multiplex_count_filtered_feature_bc_matrix.h5
wget "https://github.com/10XGenomics/HumanColonCancer_VisiumHD/raw/refs/heads/main/MetaData/SingleCell_MetaData.csv.gz"
```

Then in R, both files were loaded in and put together to create a seurat object. This is the same object that we will be loading in the next step of this lesson:

```{r}
#| eval: false
counts <- Read10X_h5("HumanColonCancer_Flex_Multiplex_count_filtered_feature_bc_matrix.h5")
meta <- read.csv("SingleCell_MetaData.csv")
rownames(meta) <- meta$Barcode

seurat <- CreateSeuratObject(counts = counts,
                             meta.data = meta,
                             project = "CRC FLEX")

# Grab UMAP coordinates and put them in a reduction
umap_coords <- as.matrix(seurat@meta.data[, c("UMAP1", "UMAP2")])

# Create a DimReduc and store it in the object as "umap"
seurat[["umap"]] <- CreateDimReducObject(embeddings = umap_coords,
                                         key = "UMAP_",
                                         assay = DefaultAssay(seurat))


# Remove cells that did not pass QC
seurat <- subset(seurat, subset = (QCFilter == "Keep"))
```

:::

So first, let us load in the reference seurat object that is provided:

```{r}
seurat_ref <- qread("data/crc_flex_ref.qs")
Idents(seurat_ref) <- "Level2"
seurat_ref <- subset(seurat_ref, 
                     downsample = 500)
seurat_ref
```

We know that there are two different levels of celltype annotations which we can begin investigating.

```{r}
#| fig-width: 20
# UMAP level 1
p1 <- DimPlot(seurat_ref, 
              group.by = "Level1", 
              label = TRUE) +
  ggtitle("CRC Reference: Level 1")

# UMAP level 2
p2 <- DimPlot(seurat_ref, 
              group.by = "Level2", 
              label = TRUE) +
  ggtitle("CRC Reference: Level 2")

p1 | p2
```

## Running RCTD

### Reference

Now that we better understand the reference, we are going to make a RCTD object out of it. This time we use the `Reference()` function and additionally supply `cluster` (celltype) information to transfer to our query. We supply the following three parameters:

- `counts` = raw counts matrix
- `cell_types` = celltype annotation
- `nUMI` = total number of counts per cell

We will be using the more specific, `Level2` annotations:

```{r}
# Raw counts of reference dataset
counts <- seurat_ref[["RNA"]]$counts

# Celltype annotation of reference dataset
cluster <- seurat_ref$Level2
cluster <- as.factor(cluster)

# Total counts of each cell
nUMI <- seurat_ref$nCount_RNA

# Create the RCTD reference object
reference <- Reference(counts = counts, 
                       cell_types = cluster, 
                       nUMI = nUMI)
```

When we look at the structure of this `reference` object, we can see that we have our `nUMI`, `counts`, and `cell_types` in this object.

```{r}
str(reference)
```

### Create query

To begin, we are going to split our seurat object into a single sample.

```{r}
nat <- subset(seurat_banksy, 
              subset = (orig.ident == "P5NAT"))
nat
```
**Deconvolution is very computationally expensive!** So we can once again make use the `sketch` workflow to downsample our dataset in a careful manner.


```{r}
# Create sketch of NAT subset
DefaultAssay(nat) <- 'Spatial.008um'

# HVGs are required input for SketchData
nat <- FindVariableFeatures(object = nat,
                            assay = "Spatial.008um")

# Calculate leverage score
nat <- SketchData(
  object = nat,
  ncells = 5000,
  method = "LeverageScore",
  sketched.assay = "sketch_nat",
  var.name = "leverage.score_nat")

# We need the PCA to later project back onto the entire dataset
nat <- FindVariableFeatures(nat)
nat <- ScaleData(nat)
nat <- RunPCA(nat, reduction.name = "pca.sketch_nat")
```

RCTD requires a unique data structure (similar to Seurat) as input to run. So we create our query object with `SpatialRNA()` by suppling the spatial coordinates and raw counts for the bins.

```{r}
DefaultAssay(nat) <- "sketch_nat"

# Grab barcodes for sketched bins
counts_sketch <- nat[["sketch_nat"]]$counts
cells_sketch <- colnames(nat[["sketch_nat"]])

# Spatial coordinates for each bin
coords <- GetTissueCoordinates(nat)[cells_sketch, 1:2]

# Create the RCTD query object
query <- SpatialRNA(coords = coords, 
                    counts = counts_sketch, 
                    nUMI = colSums(counts_sketch))
```

```{r}
str(query)
```

### Run RCTD

The more celltypes you have the longer this will take.

::: callout-warning
We need to provide a time estimate for this step

Start = 8:02
End = ~ 9:00  

This took me 
:::

```{r}
#| eval: false
# Run RCTD
RCTD <- create.RCTD(spatialRNA = query, 
                    reference = reference, 
                    max_cores = 4)
RCTD <- run.RCTD(RCTD, doublet_mode = "doublet") # This step takes a while
```

```{r}
#| echo: false
# qsave(RCTD, "intermediate/12_RCTD_nat.qs")
RCTD <- qread("intermediate/12_RCTD_nat.qs")
```

## RCTD Results

The resultant dataframe from RCTD will contain the following columns according to the [documentation](https://mirror.accum.se/mirror/bioconductor.org/packages/devel/bioc/vignettes/spacexr/inst/doc/rctd-tutorial.html):

-   `spot_class`, a factor variable representing RCTD's classification in doublet mode:
    -   "singlet" (1 cell type on pixel)
    -   "doublet_certain" (2 cell types on pixel)
    -   "doublet_uncertain" (2 cell types on pixel, but only confident of 1)
    -   "reject" (no prediction given for pixel)
-   `first_type` column gives the first cell type predicted on the bead (for all spot_class conditions except "reject").
-   `second_type` column gives the second cell type predicted on the bead for doublet spot_class conditions (not a confident prediction for "doublet_uncertain").


The results of deconvolution are stored in the `results$results_df` slot of the `RCTD` object:

```{r}
df_rctd_results <- RCTD@results$results_df
df_rctd_results %>% head()
```

We can also visualize the identifications established for each `first_type`:

```{r}
ggplot(df_rctd_results,
       aes(x = first_type, fill = first_type)) +
  geom_bar() +
  theme_bw() + ggtitle("RCTD First Type: P5NAT Sample") +
  NoLegend() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

## Project RCTD results on all bins

Let us take a look at how many rows (bins) we have these annotations for:

```{r}
nrow(df_rctd_results)
```

But didn't our `sketch` assay include 5,000 bins? RCTD runs its own QC and when it cannot identify what the annotation for a specific bin is, then it is removed from the final results.

So we are going to do some data wrangling to ensure that we keep track of all information. Including those ~800 bins without `first_type` labels.

```{r}
# Create column cell to merge on
# Also make sure columns are not factors
df_rctd_results <- df_rctd_results %>%
  mutate(cell = rownames(df_rctd_results)) %>%
  mutate(first_type = as.character(first_type),
         second_type = as.character(second_type),
         spot_class = as.character(spot_class))

# Update column names to have _sketch
# This is so that we can create a projected column
colnames(df_rctd_results) <- paste0(colnames(df_rctd_results), 
                                    "_sketch")
df_rctd_results %>% head()
```

Now let us merge this information into our Seurat object. THis is necessary step so that we can merge i

```{r}
nat$cell <- rownames(nat@meta.data)
meta <- nat@meta.data
meta <- left_join(meta, 
                  df_rctd_results,
                  by = c("cell" = "cell_sketch"))
rownames(meta) <- meta$cell
nat@meta.data <- meta

# Set value ot unassigned to sketch bins that do not have values
# This is because ProjectData will throw errors if there is not a value
# For every sketch bin
nat@meta.data[sketch_cells, "spot_class_sketch"][
  is.na(nat@meta.data[sketch_cells, "spot_class_sketch"])] <- "unassigned"
nat@meta.data[sketch_cells, "first_type_sketch"][
  is.na(nat@meta.data[sketch_cells, "first_type_sketch"])]   <- "unassigned"
nat@meta.data[sketch_cells, "second_type_sketch"][
  is.na(nat@meta.data[sketch_cells, "second_type_sketch"])] <- "unassigned"

```


```{r}
table(nat$first_type_sketch)
```

## Project Back to Full Dataset

```{r}
# Project back first_type to new column called full_first_type
crc <- ProjectData(
  object = crc,
  assay = "Spatial.016um",
  sketched.assay = "sketch_crc",
  full.reduction = "pca_crc",
  sketched.reduction = "pca.sketch_crc",
  dims = 1:30,
  refdata = list(full_first_type = "first_type",
                 full_second_type = "second_type"))
```

## Visualize New Annotation

```{r}
SpatialDimPlot(crc,
               group.by = "full_first_type")
```


***

[Next Lesson >>](tmp.qmd)

[Back to Schedule](../schedule/schedule.qmd)