---
title: "Deconvolution - Answer Key"
author:
  - Noor Sohail
date: "2026-04-19"
license: "CC-BY-4.0"
editor_options: 
  markdown: 
    wrap: 72
---

```{r}
#| label: load_libraries_data
#| echo: false
# Load libraries and data
library(Seurat)
library(tidyverse)

seurat_rctd <- qs2::qs_read("intermediate/12_seurat_RCTD.qs")
```

# Exercise 1

1. What is the sample-specific proportion for each `first_type`? Create a ggplot barplot showing the proportions of each cell type. _Hint: Refer back to the code we used [here](09_integration.qmd#fig-cell_proportion_barplot)_.
```{r}
# Barplot of proportion of cells in each first_type by sample
ggplot(seurat_rctd@meta.data) +
    geom_bar(aes(x = first_type, 
                 fill = orig.ident), 
             position = position_fill())  +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

2. Create a dotplot using the known marker genes for each celltype to see if the `first_type` labels align with known celltype genes.

```{r}
#| label: marker_list
marker_list <- list(
  "B cells" = c("IGKC", "IGHM", "CD79A", "MS4A1", "MZB1"),
  "Endothelial cells" = c("PECAM1", "VWF", "PLVAP", "ENG", "KLF2"),
  "Fibroblasts" = c("COL1A1", "COL3A1", "DCN", "LUM", "COL6A2"),
  "Intestinal epithelial cells" = c("CLCA1", "FCGBP", "MUC2", "PIGR", "ZG16"),
  "Myeloid cells" = c("C1QC", "SELENOP", "SPP1", "LYZ", "CD68"),
  "Neural cells" = c("NRXN1", "L1CAM", "NCAM1", "VIP", "CALB2"),
  "Smooth muscle cells" = c("TAGLN", "ACTA2", "MYH11", "MYL9", "CNN1"),
  "T cells" = c("TRAC", "CD3E", "TRBC2", "IL7R", "CD52"),
  "Tumor cells" = c("CEACAM6", "CEACAM5", "EPCAM", "KRT8", "LCN2")
)
```

```{r}
#| label: dotplot
#| fig-width: 15
DotPlot(seurat_rctd,
        marker_list,
        group.by = "first_type",
        cluster.idents = TRUE) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

We see that the different celltype markers do correspond quite highly with each of the `first_type` RCTD results. This gives us more confidence in the results.


***

[Back to Lesson >>](12_deconvolution.qmd)

[Back to Schedule](../schedule/schedule.qmd)