---
title: "Differential Gene Expression - Answer Key"
author:
  - Noor Sohail
date: "2026-04-29"
license: "CC-BY-4.0"
editor_options: 
  markdown: 
    wrap: 72
---

```{r}
#| label: load_libraries_data
#| echo: false
# Load libraries and data
# Load libraries and data
library(Seurat)
library(tidyverse)
library(EnhancedVolcano)

# ORA and GSEA libraries
library(tidyverse)
library(clusterProfiler)
library(org.Hs.eg.db)
library(msigdbr)

set.seed(12345)

seurat_rctd <- qs2::qs_read("intermediate/12_seurat_RCTD.qs")
```

# Exercise 1

1. Pick a different cell type and run `FindMarkers()`. Name the population you chose and print the top 6 genes in the results.

```{r}
#| label: dge_cell_types
# Vector of unique cell types
cell_types <- seurat_rctd$first_type %>%
  unique() %>% sort()

# Store DGE results
list_dge <- list()

for (ct in cell_types) {
  # Subset to cell type
  seurat_sub <- subset(seurat_rctd,
                        first_type == ct)
  Idents(seurat_sub) <- "orig.ident"

  # Calculate DGEs with wilcox test
  dge <- FindMarkers(seurat_sub,
                    ident.1 = "P5CRC",
                    ident.2 = "P5NAT")
  dge$gene <- rownames(dge)
  # Store full DGE results
  list_dge[[ct]] <- dge

  # Filter significant genes
  dge_sig <- dge %>% subset(p_val_adj < 0.05)
  # Get the gene names and get the first 6 values
  genes <- dge_sig %>%
    pull(gene) %>%
    head(6)

  # Print top 6 genes
  print(ct)
  print(genes)
}
```

# Exercise 2

2. Run an ORA analysis on the cell type that you calculated DE results for in the previous exercise. What is the top pathway that appears?

```{r}
#| label: ora_cell_types
for (ct in names(list_dge)) {
  dge <- list_dge[[ct]]

  # Background genes
  all_genes <- as.character(dge$gene)

  # Significant up-regulated genes
  sigUp <- dplyr::filter(dge,
                         p_val_adj < 0.05,
                         avg_log2FC > 0)
  sigUp_genes <- as.character(sigUp$gene)

  # GO enrichment
  egoUp <- enrichGO(gene = sigUp_genes,
                    universe = all_genes,
                    keyType = "SYMBOL",
                    OrgDb = org.Hs.eg.db,
                    ont = "BP",
                    pAdjustMethod = "BH",
                    qvalueCutoff = 0.05,
                    readable = TRUE)

  cluster_summaryUp <- data.frame(egoUp)
  print(paste(ct, ":", egoUp$Description[1]))
}
```

# Exercise 3

3. Run GSEA on the cell type that you calculated DE results for in the first exercise. What is the top pathway that appears?

```{r}
#| label: gsea_cell_types
# Use a specific collection; C5 GO signatures
m_t2g <- msigdbr(species = "Homo sapiens",
                 collection = "C5") %>%
  dplyr::select(gs_name, gene_symbol)

for (ct in names(list_dge)) {
  dge <- list_dge[[ct]]
  
  # Filter significant genes
  dge_sig <- subset(dge, p_val_adj < 0.05)
  
  # Extract fold changes
  foldchanges <- dge_sig$avg_log2FC
  
  # Name each fold change with the corresponding gene symbol
  names(foldchanges) <- dge_sig$gene
  
  # Sort fold changes in decreasing order
  foldchanges <- sort(foldchanges, decreasing = TRUE)
  
  # Run GSEA
  msig_GSEA <- GSEA(foldchanges,
                    TERM2GENE = m_t2g,
                    verbose   = FALSE)
  
  # Extract the GSEA results
  msigGSEA_results <- msig_GSEA@result
  
  # Order by NES and pull first row's ID
  top_id <- msigGSEA_results %>%
    arrange(desc(abs(NES))) %>%
    head(1) %>%
    pull(ID)
  
  print(paste(ct, ":", top_id))
}
```

***

[Back to Lesson >>](13_dge_pathway.qmd)

[Back to Schedule](../schedule/schedule.qmd)