---
title: "Spatially Variable Genes"
description: |
  Write a description of the lesson here. 
author:
  - Noor Sohail
date: "2026-03-20"
categories: 
  - category_1
  - category_2
  - category_3
  - category_4
keywords:
  - keyword_1
  - keyword_2
  - keyword_3
  - keyword_4
  - keyword_5
  - keyword_6
license: "CC-BY-4.0"
editor_options: 
  markdown: 
    wrap: 72
---

```{r}
#| label: load_libraries_data
#| echo: false
# Load libraries and data
library(Seurat)
library(tidyverse)
library(qs2)
seurat <- qs_read("intermediate/09_seurat_banksy_2.qs")
seurat
```

```{r}
# Increases the size of the default vector
# 8GB
options(future.globals.maxSize = 8000 * 1024^2)
```

Approximate time: XX minutes

## Learning objectives 

In this lesson, we will:

- Learning Objective 1
- Learning Objective 2
- Learning Objective 3

## Overview of lesson

When doing XYZ...

## Topic 1

```{r}
crc <- subset(seurat,
              orig.ident == "P5CRC")
rm(seurat)
```

```{r}
#| fig-width: 12
df <- FetchData(crc,
                c("x", "y", "seurat_cluster.projected"))

ggplot(df,
       aes(x = x, y = y,
           color = seurat_cluster.projected)) +
  geom_point(size = 0.05) +
  theme_bw() +
  facet_wrap(~seurat_cluster.projected) +
  NoLegend()
```

```{r}
crc_cluster <- subset(crc, 
              seurat_cluster.projected == 1)
crc_cluster
```



```{r}
crc_cluster <- ScaleData(crc_cluster)

test <- FindSpatiallyVariableFeatures(crc_cluster,
                                      selection.method = "moransi",
                                      verbose = TRUE,
                                      features = VariableFeatures(crc)[1:100],
                                      image = "P5CRC.008um")
test
```

```{r}
# Get spatially variable features (returns feature names)
# svf <- SpatiallyVariableFeatures(test, method = "moransi")

# Get top N features
top10_svf <- SpatiallyVariableFeatures(test, 
                                       method = "moransi")[1:10]
top10_svf
```

```{r}
#| fig-width: 12
# Spatial plot of top spatially variable features
SpatialFeaturePlot(crc_cluster,
                   features = top10_svf[1:4],
                   pt.size.factor = 15,
                   image.alpha = 0.1, 
                   ncol = 4)
```


### Subsection 1A

### Subsection 1B

:::{.callout-tip}
# [**Exercise 1**](tmp-Answer_key.qmd#exercise-1)
1. A question to evaluate Learning Objective 1
2. A followup question to question #1
3. ...
:::

## Topic 2

### Subsection 2A

### Subsection 2B

:::{.callout-tip}
# [**Exercise 2**](tmp-Answer_key.qmd#exercise-2)
1. A question to evaluate Learning Objective 2
2. A followup question to question #1
3. ...
:::

## Topic 3

### Subsection 3A

### Subsection 3B

:::{.callout-tip}
# [**Exercise 3**](tmp-Answer_key.qmd#exercise-3)
1. A question to evaluate Learning Objective 3
2. A followup question to question #1
3. ...
:::

***

[Next Lesson >>](tmp.qmd)

[Back to Schedule](../schedule/schedule.qmd)