---
title: "CellChat"
description: |
  Analyze cell–cell communication in spatial transcriptomics using CellChat infer signaling networks. You will build interaction graphs between cell types and relate spatial proximity to signaling strength.
author:
  - Noor Sohail
date: "2026-04-19"
categories: 
  - Spatial transcriptomics
  - Cell-cell communication
  - CellChat
  - Ligand-receptor analysis
keywords:
  - Ligands
  - Receptors
  - Interaction graph
  - Signaling strength
license: "CC-BY-4.0"
editor_options: 
  markdown: 
    wrap: 72
---

```{r}
#| label: load_libraries_data
#| echo: false
# Load libraries and data
library(Seurat)
library(tidyverse)
library(CellChat)

set.seed(12345)

# Set parallelization (multithreading) to speed up calculations
plan("multicore", workers = parallel::detectCores() - 1)

# Increases the size of the default vector (8GB)
options(future.globals.maxSize = 8000 * 1024^2)

seurat_rctd <- qs2::qs_read("intermediate/12_seurat_RCTD.qs")
```

Approximate time: XX minutes

## Learning objectives 

In this lesson, we will:

- Learning Objective 1
- Learning Objective 2
- Learning Objective 3

## Overview of lesson

When doing XYZ...

## Cell-cell communication


## P5CRC

We will run CellChat on one sample at a time.

```{r}
#| label: header_libraries
#| eval: false
# CellChat
# Visium HD spatial transcriptomics workshop
# Author: Harvard Chan Bioinformatics Core
# Created: December 2025

# Load libararies
library(tidyverse)
library(Seurat)
library(CellChat)
set.seed(12345)

# Set parallelization (multithreading) to speed up calculations
plan("multicore", workers = parallel::detectCores() - 1)

# Increases the size of the default vector (8GB)
options(future.globals.maxSize = 8000 * 1024^2)

# Load dataset
seurat_rctd <- readRDS("data/intermediate_seurat/seurat_RCTD.rds")
```

This is the recommended way, where you would run the full workflow on each sample and then compare and contrast.

```{r}
crc <- subset(seurat_rctd,
              subset = (orig.ident == "P5CRC") &
                       (first_type != "unassigned"))
crc
```

To create a CellChat object, we need the following information:

- Normalized counts matrix
- Spatial coordinates for each bin
- Spatial factors (stored in json file)

```{r}
# Get normalized counts matrices
counts_norm_crc <- GetAssayData(crc,
                                assay = "Spatial.008um",
                                layer = "data")
counts_norm_crc[1:5, 1:5]
```


```{r}
# Get spatial coordinates
coords_crc <- FetchData(crc, vars = c("x", "y"))
coords_crc %>% head()
```

Now we need to grab the `spatial.factors` which is defined as:

>  data frame containing two distance factors 'ratio' and 'tol', which is dependent on spatial transcriptomics technologies (and specific datasets).

```{r}
# Scale factors
sf_json <- jsonlite::fromJSON("data/P5CRC_cropped/binned_outputs/square_008um/spatial/scalefactors_json.json")
spatial_factors_crc <- data.frame(
  # pixels-to-µm conversion factor
  ratio = sf_json$bin_size_um / sf_json$spot_diameter_fullres,
  # neighbour tolerance radius in µm
  tol   = sf_json$bin_size_um / 2)
spatial_factors_crc
```


```{r}
# Create cellchat object
cellchat_crc <- createCellChat(
  object          = counts_norm_crc,
  meta            = crc@meta.data,
  group.by        = "first_type",
  datatype        = "spatial",
  coordinates     = coords_crc,
  spatial.factors = spatial_factors_crc)
```
We are working with human samples so we can make use of the human database. So we assign this to our cellchat object.

```{r}
# Load CellChat databases
cellchat_db <- CellChatDB.human
ppi <- PPI.human

# Assign DB to cellchat object
cellchat_crc@DB <- cellchat_db
```

```{r}
# Subset to signaling genes only – reduces memory and speeds up computation.
cellchat_crc <- subsetData(cellchat_crc)

# Identify genes that are over-expressed relative to other cell types
cellchat_crc <- identifyOverExpressedGenes(cellchat_crc)
cellchat_crc <- identifyOverExpressedInteractions(cellchat_crc)
```

```{r}
#| eval: false
# (Optional) Smooth expression over the protein-protein interaction network to
# reduce dropout noise. If you run this, set raw.use = FALSE below.
cellchat_crc <- smoothData(cellchat_crc, adj = PPI.human)
# cellchat <- projectData(cellchat, PPI.human) # older version
```

Here we actually run the ligand-receptor analysis:

```{r}
#| eval: false
# distance.use = TRUE  : spatial distances constrain which cell pairs can talk
# interaction.range    : max diffusion distance in µm (literature: ~250 µm)
# scale.distance       : scaling factor; tune if you get warnings about distances
# contact.dependent    : TRUE adds juxtacrine / cell-contact signalling
# contact.range        : max distance (µm) for contact-dependent signals;
#                        ~8-10 µm = cell diameter for single-cell res data,
#                        ~100 µm  = centre-to-centre distance for 10x Visium

cellchat_crc <- computeCommunProb(
  cellchat_crc,
  distance.use       = TRUE,
  interaction.range = 200,
  scale.distance     = 0.2,
  contact.dependent  = TRUE,
  contact.range      = 10,
  raw.use            = FALSE
)
```
```{r}
#| echo: false
# qs2::qs_save(cellchat_crc, "intermediate/15_cellchat_crc_computeCommunProb.qs")
cellchat_crc <- qs2::qs_read("intermediate/15_cellchat_crc_computeCommunProb.qs")
```

Now we do a final step of filtration and calculate probability of interactions occuring based upon the number and weights of the interactions:

```{r}
# removes cell-type pairs with too few cells
# and drops pathways that end up with no valid interactions
cellchat_crc <- filterCommunication(cellchat_crc, min.cells = 5)

# Summarise L-R pair probabilities into signalling pathway probabilities
cellchat_crc <- computeCommunProbPathway(cellchat_crc)
# Count interactions and sum weights into a flat network matrix
cellchat_crc <- aggregateNet(cellchat_crc)
# Compute how much each cell type sends vs receives across all pathways
cellchat_crc <- netAnalysis_computeCentrality(cellchat_crc, slot.name = "netP")
```

```{r}
#| echo: false
#qs2::qs_save(cellchat_crc, "intermediate/15_cellchat_crc.qs")
cellchat_crc <- qs2::qs_read("intermediate/15_cellchat_crc.qs")
```

We can see the structure of how data is stored within the CellChat object:

```{r}
# str(cellchat_crc)
```

### Visualizations

Now we can begin visualizing how the bins in `P5CRC` are interacting with one another:

Count + strength of interactions:

```{r}
par(mfrow = c(1, 2), xpd = TRUE)
netVisual_circle(cellchat_crc@net$count,
                 weight.scale   = TRUE,
                 label.edge     = FALSE,
                 edge.width.max = 12,
                 title.name     = "Number of interactions")
netVisual_circle(cellchat_crc@net$weight,
                 weight.scale   = TRUE,
                 label.edge     = FALSE,
                 edge.width.max = 12,
                 title.name     = "Interaction strength")
```



Heatmap of number of interactions and strength:

```{r}
gg_h_count  <- netVisual_heatmap(cellchat_crc, measure = "count",
                                  color.heatmap = "Blues",
                                  title.name    = "Number of interactions")
gg_h_weight <- netVisual_heatmap(cellchat_crc, measure = "weight",
                                  color.heatmap = "Reds",
                                  title.name    = "Interaction strength")
gg_h_count + gg_h_weight
```

Signaling scatterplot

```{r}
netAnalysis_signalingRole_scatter(cellchat_crc)
```

Incoming and outgoing heatmap for signaling strength:

```{r}
#| fig-width: 10
#| fig-height: 15
ht_out <- netAnalysis_signalingRole_heatmap(cellchat_crc, pattern = "outgoing",
                                             width         = 10,
                                             height        = 20,
                                             color.heatmap = "YlOrRd",
                                             title         = "Outgoing signaling strength")
ht_in  <- netAnalysis_signalingRole_heatmap(cellchat_crc, pattern = "incoming",
                                             width         = 10,
                                             height        = 20,
                                             color.heatmap = "GnBu",
                                             title         = "Incoming signaling strength")
ht_out + ht_in
```


Ranked barchart of interactions

```{r}
#| fig-height: 8
rankNet(cellchat_crc, 
        measure = "weight",
        mode = "single", 
        stacked = TRUE, 
        do.stat = TRUE)
```

Bubble plot ligand-receptor interactions

```{r}
#| fig-width: 15
#| fig-height: 8
n_ct <- seurat_rctd@meta.data %>%
  pull(first_type) %>%
  unique() %>% length()

# Many pathways are significant so we are going to zoom in on top 5
top_pathways <- cellchat_crc@netP$pathways[1:5]

netVisual_bubble(cellchat_crc,
                 signaling      = top_pathways,
                 sort.by.source.priority = TRUE,
                 sources.use    = 1:n_ct,
                 targets.use    = 1:n_ct,
                 remove.isolate = TRUE,
                 color.heatmap = "viridis",
                 angle.x        = 90)
```

We can also identify which ligand-receptor pairs are most significant

```{r}
df_lr <- subsetCommunication(cellchat_crc) %>%
  arrange(desc(prob))
df_lr %>% head(10)
```

As an example, we are going to look an interaction occuring between two different population (not circular):

```{r}
idx <- 7
ligand <- df_lr[idx, "ligand"]
receptor <- df_lr[idx, "receptor"]

ligand <- "COL3A1"
receptor <- "ADGRG1"

SpatialFeaturePlot(crc, 
                   c(ligand, receptor),
                   pt.size.factor = 15,
                   image.alpha = 0.2)
```

From the results, we know that this is a ligand-recptor interaction that is significant in interactions from Fibroblasts to Tumor cells. Therefore, we would expect to see higher expression of `COL3A1` in Fibroblases and `ADGRG1` in Tumor cells.

```{r}
#| fig-height: 6
VlnPlot(crc, 
        c(ligand, receptor), 
        group.by = "first_type",
        pt.size = 0,
        ncol = 1)
```

And even plot the interactions on the spatial slide


```{r}
# netVisual_aggregate on the whole pathway to shows arrows of interacttions
netVisual_aggregate(cellchat_crc,
                    signaling         = "CEACAM",
                    layout            = "spatial",
                    point.size        = 0.05,
                    alpha.image       = 0.1,
                    alpha             = 0.6,
                    vertex.size.max   = 3,
                    edge.width.max    = 5,
                    vertex.label.cex = 3)
```

```{r}
#| fig-width: 8
#| fig-height: 5
# netVisual_aggregate on the whole pathway to show chrods
netVisual_aggregate(cellchat_crc,
                    signaling  = "CEACAM",
                    layout     = "chord",
                    title.name = "CEACAM signaling — CRC")
```

## Session information

Now we would want to save the output of sessionInfo():

```{r}
#| eval: false
# Create and save a text file with sessionInfo
sink("sessionInfo_spatial_transcriptomics.txt")
sessionInfo()
sink()
```

***

[Back to Schedule](../schedule/schedule.qmd)