---
title: "Generating the Reference Dataset for RCTD"
description: |
  Write a description of the lesson here. 
author:
  - Noor Sohail
date: "2026-04-19"
categories: 
  - category_1
  - category_2
  - category_3
  - category_4
keywords:
  - keyword_1
  - keyword_2
  - keyword_3
  - keyword_4
  - keyword_5
  - keyword_6
license: "CC-BY-4.0"
editor_options: 
  markdown: 
    wrap: 72
---

```{r}
#| label: load_libraries_data
#| echo: false
# Load libraries and data

```

Approximate time: XX minutes

## Learning objectives 

In this lesson, we will:

- Learning Objective 1
- Learning Objective 2
- Learning Objective 3

## Overview of lesson

When doing XYZ...

## Download the dataset


The reference dataset was generated by downloading two files with `bash` in terminal:

- Count matrix (`.h5` file)
- Metadata csv

```{bash}
#| label: download_dataset
#| eval: false
# Download count matrix
curl -O https://cf.10xgenomics.com/samples/cell-exp/8.0.0/HumanColonCancer_Flex_Multiplex/HumanColonCancer_Flex_Multiplex_count_filtered_feature_bc_matrix.h5

# Download metadata
wget "https://github.com/10XGenomics/HumanColonCancer_VisiumHD/raw/refs/heads/main/MetaData/SingleCell_MetaData.csv.gz"
```

## Create Seurat object

Then in R, both files were loaded in and put together to create a seurat object.

```{r}
#| label: create_seurat
#| eval: false
# Load counts matrix
counts <- Read10X_h5("HumanColonCancer_Flex_Multiplex_count_filtered_feature_bc_matrix.h5")

# Load metadata and set rownames
meta <- read.csv("SingleCell_MetaData.csv")
rownames(meta) <- meta$Barcode

# Create Seurat object
seurat <- CreateSeuratObject(counts = counts,
                             meta.data = meta,
                             project = "CRC FLEX")

# Grab UMAP coordinates from metadata and put them in a reduction
umap_coords <- as.matrix(seurat@meta.data[, c("UMAP1", "UMAP2")])

# Create a DimReduc and store it in the object as "umap"
seurat[["umap"]] <- CreateDimReducObject(embeddings = umap_coords,
                                         key = "UMAP_",
                                         assay = DefaultAssay(seurat))


# Remove cells that did not pass QC
seurat <- subset(seurat, 
                 subset = (QCFilter == "Keep"))

# Set Idents to Level1 celltype annotation
Idents(seurat) <- "Level1"

# Randomly downsample dataset such that there are
# 500 cells per Level1 identity
seurat_down <- subset(seurat, 
                      downsample = 500)

saveRDS(seurat_down, "crc_flex_ref_downsample.RDS")
```

This `seurat_down` is the same object that we will be using in the [Deconvolution](12_deconvolution.qmd) lesson as the reference dataset.

***

[Back to Lesson >>](12_deconvolution.qmd)

[Back to Schedule](../schedule/schedule.qmd)